Slice Silac preprocessing
========================================================

```{r, echo=FALSE, warning=FALSE, message=FALSE}
#options(scipen=500)
suppressPackageStartupMessages(library("ggplot2"))
suppressPackageStartupMessages(library("kableExtra"))
suppressPackageStartupMessages(library(stringr))

#slices.used.for.shift <- c(21:47)
```

### Remove contaminants

Proteins annotated by MaxQuant as contaminants and reverse sequences are excluded. Additionally proteins containing one of the following expressions in the protein names are excluded:

*   Keratin
*   Desmo
*   Plako
*   Filaggrin
*   Corneao
*   Derm
*   Cystatin
*   Horn
*   TGM1

```{r, echo=FALSE, warning=FALSE, fig.width=14}

inputFileName <- "proteinGroups.txt"
outputFileName <- "output_scaled.txt"
outputFittingFunction <- "outputFittingFunction.RData"
all.tags <- c()

## open input file and load all data
dataMQ <- read.table(inputFileName, stringsAsFactors=FALSE,quote="\"", row.names=NULL,
                     header=TRUE, sep="\t", fill=TRUE, na.strings=c("Non Num\303\251rique"))


# remove columns containing this tag
if(exists("ignore.tag")){
  dataMQ <- dataMQ[,-grep(ignore.tag, colnames(dataMQ))]
}

# remove tag from colnames
for(one.tag in all.tags){
  colnames(dataMQ) <- gsub(one.tag, "", colnames(dataMQ))  
}

# remove contaminants identified by their names
contaminant.term <- c("keratin", "desmo", "plako", "filaggrin", "corneao", "derm", "cystatin", "horn", "annexin")
contaminant.ac <- c("Q9NZT1", "P31944", "P04264")
contaminant.id <- grep( paste("^", contaminant.term, sep="", collapse="|"), dataMQ[, "Fasta.headers"], ignore.case = TRUE)
contaminant.id <- c(contaminant.id, grep( paste(contaminant.ac, sep="", collapse="|"), dataMQ[, "Protein.IDs"]))

# remove given gene names
contaminant.genes <- c("TGM1")
contaminant.id <- c(contaminant.id, grep( paste("^", contaminant.genes, sep="", collapse="|"), dataMQ[, "Fasta.headers"], ignore.case = TRUE))
contaminant.id <- unique(contaminant.id)

# prepare table to plot
contaminants.removed <- dataMQ[contaminant.id, c("Protein.IDs", "Fasta.headers", "Potential.contaminant", "Ratio.H.L")]
contaminants.removed$Protein.IDs <- unlist(lapply(contaminants.removed$Protein.IDs, function(x) head(unlist(strsplit(x, ";")) , n=1)))

# remove also contaminants and reverse indicated by MaxQuant
contaminant.id <- c(contaminant.id, which(dataMQ$Only.identified.by.site == "+"))
contaminant.id <- c(contaminant.id, which(dataMQ$Reverse == "+"))
contaminant.id <- c(contaminant.id, which(dataMQ$Potential.contaminant == "+"))

contaminant.id <- unique(contaminant.id)
dataMQ <- dataMQ[-contaminant.id, ]

```

#### Contaminants removed because of their names

```{r,results='asis', echo=FALSE}
#table <- gvisTable(contaminants.removed, options=list(width=1200, height=450))
#print(table, tag="chart")
contaminants.removed %>%
  kable() %>%
  kable_styling(c("striped", "condensed"), full_width = FALSE)
```

```{r, echo=FALSE, warning=FALSE, fig.width=14}
# get the columns of interest
data.colnames <- colnames(dataMQ)
countColumns <- grep("Ratio.H.L.count\\.[0-9]?\\.?\\w+", data.colnames)
band.names <- sub("Ratio.H.L.count.", "", data.colnames[countColumns])
ratioColumns <- which(data.colnames %in% paste0("Ratio.H.L.", band.names))

# get the rigth slice numbering
slices_list <- unlist(lapply(colnames(dataMQ[,ratioColumns]), function(x) sprintf("%02d",as.numeric(str_extract(x, "([0-9]+)$")))))
slices_number <- length(slices_list)

## keep only high expressed
isAbove10 <- apply(dataMQ[, ratioColumns ], 
                   MARGIN=1, 
#                   FUN=function(x) {return(any(as.numeric(x)>10, na.rm=TRUE))})
                   FUN=function(x) {return(FALSE)})

dataMQ <- dataMQ[!isAbove10, ]

## Order by theoretical molecular weight
molWeightColumn <- "Mol..weight..kDa."
protOrderByWeight <- order(as.numeric(dataMQ[, molWeightColumn]), decreasing=TRUE)

sortedDataMQ <- dataMQ[protOrderByWeight, countColumns]
sortedMolWeight <- as.numeric(dataMQ[protOrderByWeight, molWeightColumn])
sortedRatios <- dataMQ[protOrderByWeight, ratioColumns]

## Boxplot for all band
whichDetected <- (sortedDataMQ > 4)
allWeightsByBand <- data.frame(matrix(rep(sortedMolWeight, times=slices_number), ncol=slices_number))
allWeightsByBand[!whichDetected] <- NA
names(allWeightsByBand) <- slices_list

stackedMolWeight <- stack(allWeightsByBand)
stackedMSMSCount <- stack(sortedDataMQ)
stackedRatios <- stack(sortedRatios)

orderByMSMSCount <- order(stackedMSMSCount$values, decreasing=FALSE)
sortedStackedMolWeight <- stackedMolWeight[orderByMSMSCount,]
sortedStackedMSMSCount <- stackedMSMSCount[orderByMSMSCount,]
sortedStackedRatios <- stackedRatios[orderByMSMSCount,]

sortedStackedRatios$ind <- sortedStackedMolWeight$ind
colnames(sortedRatios) <- slices_list


# prepare ticks for log2 scale:
log.ticks <- c(12.5)
end.tick <- 800

do.ticks <- function(x){
  last.x <- x[length(x)]
  if(last.x < end.tick) x <- Recall(c(x, last.x*2))
  x
}

log.ticks <- do.ticks(log.ticks)
```

### Protein intensitiy plot

This plot shows the intensities of the proteins (H/L ratios count).
The x-axis represents the corresponding slices, the y-axis the theoretical molecular weight of the protein identification.
Only proteins with an MSMS count > 4 are shown.

```{r, echo=FALSE, warning=FALSE, fig.width=14}

# the plot
qplot(x=ind, y=values, data=sortedStackedMolWeight, na.rm=TRUE, geom=c("jitter"), 
      colour=log2(sortedStackedMSMSCount$values), 
      xlab="Slice", ylab="Theoretical molecular weight (log2)") + 
    scale_colour_gradientn(colours=c("yellow", "red"), name="H/L ratios count (log2)") + 
    scale_y_continuous(breaks=log.ticks) + 
    coord_trans(y = "log2") 
```

### Protein ratio plot

This plot shows the H/L ratios.
The x-axis represents again the corresponding slices, the y-axis the theoretical molecular weight of the protein identification.
Only proteins with an MSMS count > 4 are shown.

```{r, echo=FALSE, warning=FALSE, fig.width=14}
qplot(x=ind, y=values, data=sortedStackedMolWeight, na.rm=TRUE, geom=c("jitter"), 
      colour=log2(sortedStackedRatios[,"values"]), 
      xlab="Slice", ylab="Theoretical molecular weight (log2)") + 
    scale_colour_gradient(low="green", high="red", name="H/L ratios (log2)", limits=c(-5.5,5.5)) +
    scale_y_continuous(breaks=log.ticks) + 
    coord_trans(y = "log2")
```

```{r, echo=FALSE, warning=FALSE}

# select either normalized or raw data ("Ratio.H.L.normalized.")
#interesting_colnames_ratios <- grep("Ratio.H.L\\.[0-9]*\\.", colnames(dataMQ)) 
interesting_colnames_ratios <- ratioColumns # grep("Ratio.H.L\\.normalized\\..+", colnames(dataMQ)) 
my.ratios <- dataMQ[, interesting_colnames_ratios]
colnames(my.ratios) <- slices_list
my.stacked.ratios <- stack(my.ratios)

# plot median of raw ratios
medianRawRatios <- apply(log2(my.ratios), 2, function(x) median(x, na.rm=TRUE))
names(medianRawRatios) <- slices_list

# zero median normalization
zero.median.norm <- apply(log2(my.ratios), 2, function(x) x-median(x, na.rm=TRUE))

# mean between slices 6 - 38
#median.correction <- round(mean(medianRawRatios[slices.used.for.shift]), digits=2)

# correct slice 4 and 37 according to their neighbours
#neighbour.correction <- median(unlist(log2(my.ratios)[c(3,5)]), na.rm=TRUE)
#zero.median.norm[,4] <- unlist(log2(my.ratios)[4] - neighbour.correction)

# the limits for the boxplots:
bp.lower.lim <- log2(min(my.ratios, na.rm=TRUE)) - 0.5
bp.upper.lim <- log2(max(my.ratios, na.rm=TRUE)) + 0.5
bp.limits <- c(bp.lower.lim, bp.upper.lim)

```

### Normalization

Non normalized data *(Ratio.H.L)* from the MaxQuant results were taken. We normalize the those data using a zero median normalization.

#### Boxplot before zero median normalization

```{r, echo=FALSE, warning=FALSE, fig.width=14}

## Not norm + log2
ggplot(my.stacked.ratios, aes(x=ind, y=log2(values)), na.rm=TRUE) + 
  geom_boxplot() +
  xlab("Slice") +
  ylab("H/L ratios (log2)") +
  ylim(bp.limits)

# ## MaxQuant normalization + log2
# interesting_colnames_ratios_MQnorm <- paste("Ratio.H.L.normalized.", slices_list, sep="") 
# stackedDataMQNorm <- stack(log2(dataMQ[!isAbove10, interesting_colnames_ratios_MQnorm]))
# boxplot(values ~ind, data=stackedDataMQNorm, names=slices_list, main="MaxQuant normalization (log2)")
# 
# 
# ## not norm + 0.25
# interesting_colnames_ratios_MQnorm <- paste("Ratio.H.L.normalized.", slices_list, sep="") 
# stackedDataMQNorm <- stack(log2(dataMQ[!isAbove10, interesting_colnames_ratios_MQnorm]))
# boxplot(values ~ind, data=stackedDataMQNorm, names=slices_list, main="MaxQuant normalization (log2)")

```


#### Boxplot after zero median normalization

All Slices were shifted to zero median.

```{r, echo=FALSE, warning=FALSE, fig.width=14}

 my.normRatios.log <- data.frame(zero.median.norm)
 colnames(my.normRatios.log) <- slices_list

 ggplot(stack(my.normRatios.log), aes(x=ind, y=values), na.rm=TRUE) + 
  geom_boxplot() +
  xlab("Slice") +
  ylab("H/L ratios (log2)") +
  ylim(bp.limits) 

```


```{r, echo=FALSE}

colnames(my.normRatios.log) <- paste0("Ratio.H.L.PAF.normalized.", colnames(my.normRatios.log))
final.dataMQ <- cbind(dataMQ, my.normRatios.log)

write.table(x=final.dataMQ, file=outputFileName, sep="\t", na='', quote=FALSE, row.names=FALSE)

#colnames(head(dataMQ))

# in case we want the MQ normalized and median-shifter ratios:
#data.for.results <- scaledVals

# in case we want the MQ normalized ratios:
# data.for.results <- log2(dataMQ[!isAbove10, interesting_colnames_ratios_MQnorm])

# in case we want the raw ratios (with a median shift of + 0.08):
# data.for.results <- log2(dataMQ[!isAbove10, interesting_colnames_ratios]) + 0.08 
# colnames(data.for.results) <- colnames(scaledVals)
# 
# ## Write table
# write.table(x=cbind(dataMQ[!isAbove10, !(names(dataMQ) %in% paste("Ratio.H.L.PAF.normalized.", slices_list, sep=""))], 
#                     data.for.results), 
#             file=outputFileName, sep="\t", na='', quote=FALSE, row.names=FALSE)

```


### Fit slices to masses

##### Plot data before filtering:

```{r, echo=FALSE, warning=FALSE, fig.width=14}

# use the functions from preprocessDataFunctions.R
source("preprocessDataFunctions.R")
library(reshape2)

# load data
dataMQList <- list()

dataMQList[["data"]] <- dataMQ
my.filename <- gsub(".txt", "", inputFileName)
dataMQList[["filename"]] <- my.filename
dataMQList[["slices_number"]] <- slices_number

# remove contaminants from data
dataMQList <- remove_MQ_contaminants(dataMQList)

# extract intensities (add "intensities" and "masses")
#ignore.slices <- c(1)
#dataMQList <- get_intensity_and_mass_columns(dataMQList, ignore.slices)
dataMQList <- get_intensity_and_mass_columns(dataMQList)

# keep the mass min/max for comparable plotting
my.ylim <- c(min(log10(dataMQList$masses)), max(log10(dataMQList$masses)))

# plot original data before filtering
p <- plot_original_data(dataMQList) + scale_y_continuous(limits=my.ylim)
print(p)

```


##### Remove points with H/L counts below 4:

```{r, echo=FALSE, warning=FALSE, fig.width=14}
dataMQList <- filter_by_H.L.count(dataMQList, 4)

p <- plot_original_data(dataMQList) + scale_y_continuous(limits=my.ylim)
print(p)
```

##### After removing proteins which are present on more than 30% of the slices (probably contaminants):

```{r, echo=FALSE, warning=FALSE, fig.width=14}

# filter entries which are repeated in to many slices. They are probably also contaminants.
dataMQList <- filter_repeated_entries(dataMQList, totPercThreshold=0.2)

# plot original data after filtering
p <- plot_original_data(dataMQList) + scale_y_continuous(limits=my.ylim)
print(p)

```

##### Remove regions with only very few proteins (< 5) and fit this data using a polynomial function:

```{r, echo=FALSE, warning=FALSE, fig.width=14}

# keep only regions with many points
dataMQList <- filter_low_densities(dataMQList, minNrThreshold=5)

# plot data
#p <- plot_original_data(dataMQList) + scale_y_continuous(limits=my.ylim)
#print(p)

# fit slices to log10(theoretical_masses)
my.fitting.function <- get_fitting_function(dataMQList)
predicted_masses <- predict_slice_masses(dataMQList, my.fitting.function)

save(my.fitting.function, predicted_masses, file=outputFittingFunction)

p <- plot_fit_data(dataMQList, my.fitting.function)
print(p)

```

##### Quality of matching and a plot showing the final fit:

```{r, echo=FALSE, warning=FALSE, fig.width=14}

#print(my.fitting.function)
summary(my.fitting.function)

predicted_masses.flat <- 10^predicted_masses
predicted_masses.stacked <- stack(predicted_masses.flat)
predicted_masses.stacked$ind = 1:(length(slices_list)) #as.factor(slices_list)

p <- qplot(x=ind, y=values, data=sortedStackedMolWeight, na.rm=TRUE, geom=c("jitter"), 
    colour=log2(sortedStackedMSMSCount$values), 
    xlab="Slice", ylab="Theoretical molecular weight (log2)") + 
    scale_colour_gradientn(colours=c("yellow", "red"), name="H/L ratios count (log2)") + 
    scale_y_continuous(breaks=log.ticks) + 
    coord_trans(y = "log2")+
    geom_line(data=predicted_masses.stacked, aes(x=ind, y=values), colour="red", group="polynomial fit", size=1.5)

print(p)


my.c <- coef(my.fitting.function)

```

##### Polynomial function used for fitting (x: slice_number, f(x): theoretical MW in log10 ):

f(x) = `r sprintf("%e", my.c[1])` + `r sprintf("%e", my.c[2])` * x  + `r sprintf("%e", my.c[3])` * x^2 + `r sprintf("%e", my.c[4])` * x^3 

##### Predicted MW's for all the slices
```{r,results='asis', echo=FALSE}
predicted_masses.table <- data.frame(slice=names(predicted_masses), theoMW=10^predicted_masses)
predicted_masses.table %>%
  kable() %>%
  kable_styling(c("striped", "condensed"), full_width = FALSE)
#table <- gvisTable(predicted_masses.table)#, options=list(width=1200, height=450))
#print(table, tag="chart")
```